---
title: "Airline"
author: "Donovan Johnson"
output:
  html_notebook: default
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

### {.tabset}

#### Packages and Meta Data

__Install and Initialize Packages__
```{r}
packages <- c("rlang" #provides various interfaces for working with R and R objects.
              ,"ggplot2" #used for data visualization  
              , "dplyr" #used to manipulate data
              , "tidyr"#used to create tidy data
              ,"readr" #used to read in data
              ,"DataExplorer" #used to do Exploratory Data Analysis and structure of data set
              , "DT" #Create Nice Data Tables
              , "ggpubr" #Arrange ggplots
              ,"ggcorrplot" #Make Correlation Plot Visualization 
              ,"viridis" #Adds more eye popping colors
              ,"skimr") #Make summary statistic tables

# Install packages not yet installed
installed_packages <- packages %in% rownames(installed.packages())
if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages])
}
# Packages loading
invisible(lapply(packages, library, character.only = TRUE))
```

__Meta Data__
```{r}
#Create a table for Meta Data
meta_data<-data.frame(Dataset=c("flights","flights","finaldf","finaldf",
                                "finaldf","finaldf","finaldf","finaldf",
                                "finaldf","finaldf","finaldf","finaldf",
                                "finaldf","agg_tbl4","agg_tbl4","agg_tbl4"),
  Field_name=c("ORIGIN_AIRPORTSIZE","DESTINATION_AIRPORTSIZE",
               "DEP15","ARR15","CPM","OOC","DOC",
               "COST","TREV","BREV","REV","PROFIT","Pair",
               "DDP","PM","Breakeven"),
  Description=c("The type of the origin airport:large or medium",
                "The type of the destination airport:large or medium",
                "Departure Delay Charge","Arrival Delay Charge",
                "Cost per mile of trip",
                "Operation Cost for Orgigin airport",
                "Operation Cost for Destination airport",
                "Cost of the Round Trip",
                "Revenue from ticket sales",
                "Revenue from bag fee",
                "Total Revenue",
                "Toal Profit",
                "Order Pair of Destination and Origin",
                "Departure Delay Per Passenger",
                "Profit Margin",
                "Amount of flights needed to breakeven"
                ))

datatable(meta_data,class='cell-border stripe',caption="Metadata",autoHideNavigation = getOption("DT.autoHideNavigation", TRUE))
```

__Import Data sets__
```{r}
flights<-read_csv("C:/Users/djrem/Desktop/rPractice/Flights.csv")
airport_codes<-read_csv("C:/Users/djrem/Desktop/rPractice/Airport_Codes.csv")
tickets<-read_csv("C:/Users/djrem/Desktop/rPractice/Tickets.csv")
```

#### Quality Check
```{r}
#Filter for only medium or large airports and only the US destinations
airport_codes <- airport_codes %>%
  filter(TYPE %in% c("medium_airport", "large_airport") & ISO_COUNTRY == "US")

#filter out trips that are not round trip
tickets <-filter(tickets,ROUNDTRIP==1)

#Remove cancelled flights from Data set
flights <- flights %>%
  filter(CANCELLED==0) 
```


```{r}
#Create Function to Look at Structure and Exploratory Data Analysis
Structure_EDA <-function(df){
#shows the structure of each column in a data set
  str(df)
  
#plots the percentage of data missing by column
plot_missing(df, geom_label_args = list("size" = 2, "label.padding" = unit(0.2, "lines")),title="Percentage of Data Missing by Column",theme_config = list(legend.position = c("none")))

#plots a histogram for each int column
plot_histogram(df,title="Distribution of Integer Data",theme_config=list(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),plot.title=element_text(size=5)))

#plots a bar graph for columns that have integer data and can be grouped realistically
plot_bar(df,title="Grouped Data or Each Column",theme_config = list(axis.text.y = element_text(size = 5),axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)))

}
```


```{r}
#Create Functions to Detect and Remove Outleirs from Numeric Columns
#Source: https://www.geeksforgeeks.org/how-to-remove-outliers-from-multiple-columns-in-r-dataframe/

# create detect outlier function
detect_outlier <- function(x) {
   
    # calculate first quantile
    Quantile1 <- quantile(x, probs=.25)
   
    # calculate third quantile
    Quantile3 <- quantile(x, probs=.75)
   
    # calculate inter quartile range
    IQR = Quantile3-Quantile1
   
    # return true or false
    x > Quantile3 + (IQR*1.5) | x < Quantile1 - (IQR*1.5)
}
 
# create remove outlier function
remove_outlier <- function(dataframe,
                            columns=names(dataframe)) {
   
    # for loop to traverse in columns vector
    for (col in columns) {
       
        # remove observation if it satisfies outlier function
        dataframe <- dataframe[!detect_outlier(dataframe[[col]]), ]
    }
   
    # return dataframe
    print("Remove outliers")
    print(dataframe)
}
```

```{r}
#Structure and EDA for Flights before transformation
Structure_EDA(flights)
```

```{r}
#AIR_TIME and DISTANCE need to be changed to numerical data
flights <- flights %>%
  mutate(
    AIR_TIME = as.numeric(AIR_TIME),
    DISTANCE = as.numeric(DISTANCE)
  )
```

```{r}
#Impute NA's using median in order to have a complete data set
flights<- flights%>% mutate(across(where(is.numeric), ~replace_na(., median(., na.rm=TRUE))))
```

```{r}
#There seems to be outliers in AIR_TIME,ARR_DELAY,DEP_DELAY, and DISTANCE and need to be identified and removed. 
flights<-remove_outlier(flights,c("AIR_TIME","ARR_DELAY","DEP_DELAY","DISTANCE"))
```


```{r}
#Structure and EDA for flights after transformation
Structure_EDA(flights)
```

```{r}
#Structure and EDA for airport_codes before transformation
Structure_EDA(airport_codes)
```


```{r}
#As Continent Column is entirely Missing it will be removed entirely
airport_codes<-select(airport_codes,c(-CONTINENT))

#As this imputing NA's is not a valid solution for this data set as the majority of the missing data set is not numeric, I will be remove all NA's

airport_codes<-na.omit(airport_codes)

```

```{r}
#Structure and EDA for airport_codes after transformation
Structure_EDA(airport_codes)
```


```{r}
#Structure and EDA for tickets before transformation
Structure_EDA(tickets)
```

```{r}
#ITIN_FARE need to be changed to numerical data
tickets <- tickets %>%
  mutate(
    ITIN_FARE = as.numeric(ITIN_FARE))
```

```{r}
#Impute NA's using median in order to have a complete data set
tickets<- tickets%>% mutate(across(where(is.numeric),
                                   ~replace_na(., median(., na.rm=TRUE))))
```

```{r}
#There seems to be outliers in ITIN_FARE and PASENGERS need to be identified and removed. 
tickets<-remove_outlier(tickets,c("ITIN_FARE","PASSENGERS"))
Structure_EDA(tickets)
```

#### Data Munging

```{r}
create_final_df <- function(tickets, flights, airport_codes) {
  # Creates a vector containing the IATA codes of medium airports
  medium_airports <- airport_codes %>%
    filter(TYPE == "medium_airport") %>%
    pull(IATA_CODE)

  # Filters out airports that are not Medium or Large
  flights_filtered <- flights %>% 
    filter(ORIGIN %in% airport_codes$IATA_CODE & 
             DESTINATION %in% airport_codes$IATA_CODE) %>%
    # Creates two columns called ORIGIN_AIRPORTSIZE and DESTINATION_AIRPORTSIZE
    mutate(
      ORIGIN_AIRPORTSIZE = if_else(ORIGIN %in% medium_airports,
                                   "medium_airport", "large_airport"),
      DESTINATION_AIRPORTSIZE = if_else(DESTINATION %in% medium_airports,
                                        "medium_airport", "large_airport"))

  # Aggregate Ticket data set to ORIGIN & DESTINATION by mean of ITIN_FARE to perform an inner_join
  agg_tbl <- tickets %>% group_by(ORIGIN, DESTINATION) %>% 
    summarise(ITIN_FARE = round(mean(ITIN_FARE), 2), .groups = 'drop') 

  # Create final data frame by performing inner join
  finaldf <- flights_filtered %>% inner_join(agg_tbl,
                                              by = c("ORIGIN" = "ORIGIN",
                                                     "DESTINATION" = "DESTINATION"))

  return(finaldf)
}

# Call the function and store the result in final_df
finaldf <- create_final_df(tickets, flights, airport_codes)
```

```{r}
#Makes a variable that makes round trips ordered one way in order to be aggregated
finaldf<- finaldf %>% mutate(Pair = ifelse(DESTINATION < ORIGIN,
                         paste(DESTINATION, ORIGIN, sep = ','),
                         paste(ORIGIN, DESTINATION, sep = ',')))
```

```{r}
head(finaldf)
```

#### Craft a Visual Data Narrative

__1.The 10 Busiest round trip routes__

```{r}
# Function to create a summarized plot
create_summary_plot <- function(data, x_var, y_var, fill_color, title, y_label) {
  # Capture y_var as a symbol using rlang's ensym() to handle variable scoping
  y_var <- ensym(y_var)
  # Summarize the data and create a bar plot
  data %>%
    top_n(n = 10, wt = !!y_var) %>%
    ggplot(aes(x = reorder(.data[[x_var]], -!!y_var), y = !!y_var)) +
    geom_bar(stat = "identity", width = 0.5, fill = fill_color) +
    # Styling
    theme_bw() +
    labs(title = title,
         x = "Flight Route",
         y = y_label) +
    theme(axis.text.x = element_text(angle = 65, vjust = 0.6)) +
    scale_y_continuous(labels = scales::comma)
}
```

```{r}
agg_tbl2 <- finaldf %>%
  group_by(Pair) %>%
  summarise(Total_Count = sum(n()), .groups = 'drop') %>%
  arrange(desc(Total_Count)) 

#Data table in order
datatable(head(agg_tbl2,10))

#Create Visualization of top 10 Busiest Round Trips
create_summary_plot(agg_tbl2, "Pair", Total_Count, "tomato3", "Ordered Bar Chart 10 Busiest Round Trip Routes", "Number of Flights")
```

__2.Most Profitable Airlines__
```{r}
##Creating Variables related to the Cost calculation and then create the cost calculation
finaldf <- finaldf %>%
  mutate(
    #Create a variable to Check if a Delay is more than 15 mins to calculate the amount owed for DEP_DELAY and ARR_DELAY
    DEP15 = pmax((DEP_DELAY - 15) * 75, 0),
    ARR15 = pmax((ARR_DELAY - 15) * 75, 0),
    #Create a variable to check amount per mile for Fuel, Oil, Maintenance, and Depreciation
    CPM = round(9.10 * DISTANCE,2),
    #Create two variables to calculate operation cost for ORIGIN and DESTINATION airports
    OOC = ifelse(ORIGIN_AIRPORTSIZE == "medium_airport", 5000, 10000),
    DOC = ifelse(DESTINATION_AIRPORTSIZE == "medium_airport", 5000, 10000), 
    COST = round((DEP15+ARR15+OOC+DOC+CPM),2)
  )
```

```{r}
#Creating variables related to the revenue calculation and then create the revenue and profit calculation
finaldf <- finaldf %>%
  mutate(
    #Create a variable to Check Ticket Revenue
    TREV = (200*OCCUPANCY_RATE*ITIN_FARE),
    #Create a variable to check bag fee revenue
    BREV = (200*OCCUPANCY_RATE*.50*70),
    #Create a variable to check Total revenue
    REV = (TREV+BREV),
    #Make profit variable
    PROFIT = round(REV-COST,2)
  )
```


```{r}
#Get the Top 10 most profitable flights grouped by pair
agg_tbl3 <- finaldf %>%
  group_by(Pair) %>%
  summarise(Sum_Profit = round(sum(PROFIT), 2),
            Sum_Cost = round(sum(COST), 2),
            Sum_Rev = round(sum(REV), 2),
            Occupancy = round(mean(OCCUPANCY_RATE), 2),
            Itin_Fare = round(mean(ITIN_FARE), 2),
            Distance = round(mean(DISTANCE), 2),
            Arrival_Delay = round(mean(ARR_DELAY), 2),
            Departure_Delay = round(mean(DEP_DELAY), 2),
            .groups = 'drop') %>%
  arrange(desc(Sum_Profit)) 

#Data table in order
datatable(head(agg_tbl3,10))

# Create a summarized plot for the 10 most profitable trips
create_summary_plot(agg_tbl3, "Pair", Sum_Profit, "blue3", "Ordered Bar Chart 10 Most Profitable Trips", "Profit in Dollars")

# Create a summarized plot for the 10 most profitable trips ordered by revenue
A<-create_summary_plot(agg_tbl3, "Pair", Sum_Rev, "green3", "10 Most Profitable Trips Ordered by Revenue", "Revenue in Dollars")

# Create a summarized plot for the 10 most profitable trips ordered by cost
B<-create_summary_plot(agg_tbl3, "Pair", Sum_Cost, "tomato3", "10 Most Profitable Trips Ordered by Cost", "Cost in Dollars")

#Compare costs and Revenue Side by Side
ggarrange(A,B,nrow=2)

#Summary statistics of Data Set
no_hist <- skim_with(ts = sfl(line_graph = NULL))
agg_tbl3 %>% no_hist() %>% yank("numeric") %>% 
  select(-c("n_missing","complete_rate","sd")) %>%datatable()

corr <- (cor(select(agg_tbl3,-c("Pair")))) 

#Create Correlogram to look at relationships between key components
ggcorrplot(corr, hc.order = TRUE, 
           type = "lower", 
           lab = TRUE, 
           lab_size = 3, 
           method="square", 
           colors = c("tomato2", "white", "springgreen3"), 
           title="Correlogram of Key Components", 
           ggtheme=theme_bw())
```

__3. My personal five best__
```{r}
agg_tbl4 <- finaldf %>%
  group_by(Pair) %>%
  summarise(Sum_Profit = round(sum(PROFIT), 2),
            Sum_Cost = round(sum(COST), 2),
            Sum_Rev = round(sum(REV), 2),
            Occupancy = round(mean(OCCUPANCY_RATE), 2),
            Itin_Fare = round(mean(ITIN_FARE), 2),
            Distance = round(mean(DISTANCE), 2),
            Arrival_Delay = round(mean(ARR_DELAY), 2),
            Departure_Delay = round(mean(DEP_DELAY), 2),
            Total_Count = sum(n()),
            .groups = 'drop') %>%
  #Create variable to calculate Departure Delay per Passenger 
  mutate(DDP=round(Departure_Delay/Occupancy*200,2),
  #Create variable to calculate to create a Pseudo Net Profit Margin formula as we are missing Interest and Taxes
         PM=(round(Sum_Profit/Sum_Rev*100,2)))
```

```{r}
#find the 75th percentile for Profit 
Q1<-quantile(agg_tbl4$Sum_Profit, probs = .95)

#finding the 90th percentile for Profit Margin
Q2<-quantile(agg_tbl4$PM, probs = .9)
```

```{r}
#Filter out SUM_Profit that are lower than the 95 percentile for Profit and 90 Percentile for Profit Margin

agg_tbl4 <- agg_tbl4 %>%
  filter(Sum_Profit >= Q1,
         PM >= Q2) %>%
#arrange by Departure Delay by Passenger
        arrange(DDP) %>% head(5)

datatable(head(agg_tbl4,10))

# Create a summarized plot for the 10 most profitable trips ordered by Departure Delay by Passenger
ggplot(data=agg_tbl4,aes(x=reorder(Pair,DDP),y=Sum_Profit,fill=PM))+ geom_bar(stat='identity')+ 
  scale_fill_viridis(option = "C")+
  labs(x="Flight Route",
       y="Total Profit",
       title="5 Recommended Flights Routes Ordered by\nDeparture Delay Per Passenger",
       fill="Profit Margin") + scale_y_continuous(labels = scales::comma)
```

__4.Number of round trips flight it will take to break even on the upfront cost for each of the 5 round trips__

```{r}
agg_tbl4<-agg_tbl4 %>% 
  #Create Break even variable
  mutate(Breakeven=90000000/(Sum_Profit/Total_Count))

datatable(agg_tbl4)

create_summary_plot(agg_tbl4, "Pair", Breakeven, "tomato3", "Amount of Flights Required to Breakeven\nfor Recommended Flights", "Amount of Flights")

#Summary Statistics for Recommended Flights
agg_tbl4 %>% no_hist() %>% yank("numeric") %>% 
  select(-c("n_missing","complete_rate","sd")) %>%datatable()
```

__5.KPI to check for recommended flights__

Considering that the company's motto is "On time, for you." I selected KPIs that emphasize punctuality and client happiness.

Considering that the company’s motto is “On time, for you.” I chose KPIs that emphasize punctuality and client happiness.

1. Delay Reason: I want to keep track of the causes of delays because it might help us understand when a certain flight route is experiencing frequent delays and how we might be able to address these issues for on-time plane departures.

2. Cancellation Rate: Like Delay Reason, I want to know when flights are being canceled, understand why they are being cancelled so that I can reduce the frequency of flight cancellation.

3. Arrival Punctuality: Displays the proportion of flights that were late. Important because we aim to maintain a high level of passenger satisfaction by maintaining flight on-time performance.

4.  Passenger Quota: Calculates the proportion of all passengers who arrived on time at the destination airport.

5. Jet Fuel Price Index: as a Percentage of Cost: Jet fuel is one of the significant expenses any airline faces. The airline does not control the fuel prices and can be highly volatile due to its dependency on several external factors. Fuel can account for anywhere between 30% - 60% of the airline's operating costs.



#### Final Recommendations:
The five round trip flights I recommend are:
1. (DFW,XNA) 2. (CID,ORD) 3. (IAH,MAF) 4. (CLT,ROA) 5. (CLT,MYR)

I chose these round-trip flights based on three important metrics. Profit, Profit margin, and Departure Delay per Passenger. The first two KPI are expected. The client shouldn't enter the US market if the round-trip flight isn't profitable or doesn't have a good margin to start with. The third metric is to emphasis the clients' priority on delivering passenger to their destination on time. Given that the client beliefs they have a competitive advantage, a metric that prioritizes their flight routes by departure delay seems paramount.


#### What's Next
First off, given that we have such a sizable and reliable data collection, I think it would be feasible to establish a test and training set and utilize supervised machine learning techniques on this data set to determine what the major influences are on DEP_Delay and ARR_Delay, Profit and Profitability.

Second, I might consider including geographic visualizations to see if I can gain additional insight into certain groups of round-trip flights and determine whether geographical regions are more lucrative than others.

Thirdly, to determine what variables in our data most strongly correlate with cancelled flights, I would also examine associations in the cancelled flight data set.

Fourthly, I would like to use machine learning techniques to impute missing data if I had more time. I choose to utilize a median imputation because this is an extremely computationally costly task.

Finally, I have the option to utilize the 95th percentile for profit and the 90th percentile for profit margin. However, rather than arranging them in ascending order, I can arrange them in descending order based on the departure delay per passenger. This approach aims to focus on markets where flights have a high profit/profit margin per roundtrip but will sort our recommendation by routes that have the most delays.


